---
title: "Tutorial"
author: "Hubert Ruczyński"
output: 
  html_document: 
    df_print: paged
    toc: true
    toc_float: true
    number_sections: true      
vignette: >
  %\VignetteIndexEntry{tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  
---
# FairPAN

In this tutorial you will get to know when, why and how to use `FairPAN`. 
`FairPAN` is a tool for creating fair predictive adversarial networks. 


## How does it work?

## Why?

## Data

The dataset used in our tutorial is called adult. It contains 15 columns with
both numerical and categorical data about citizens of USA. Our *target* here is
a `salary` column. As *sensitive* variables we can perceive `sex` and `race`
however we will focus only on `sex`

```{r}
library(FairPAN)
adult <- fairmodels::adult
head(adult)
```

# Workflow (?)

## Preprocessing

At the beginning we have to preprocess our dataset so we can train a neural
network on it and divide it into train and test subsets. One can do it on your
own, however we will use built in function called preprocess which will create
16 objects which are needed for other features.

To use this function we have to provide a dataset with categorical columns 
provided as factors. Then we define that `salary` is our target and `sex` is a
sensitive variable with privileged level `Male` and discriminated `Female`. 
As we noticed before, `race` could also be considered as a sensitive variable
so we want to remove it from our learning dataset too. In the end we sample a
small part of the dataset for our process to save our time in this example and
define proportions of train, test and validation subsets. We also set seed for
reproduction.

Inside, the function encodes the categorical columns as integers based on their
factors levels. After that all variables are rescaled to ensure better learning
process. Moreover to ensure that adversarial model works properly, we also 
balance a dataset to have the same number of privileged and discriminated
records.

```{r}
data <- preprocess( data = adult,               # dataset 
                    target_name = "salary",     # name of target column
                    sensitive_name = "sex",     # name of sensitive column
                    privileged = "Male",        # level of privileged class
                    discriminated = "Female",   # level of discriminated class
                    drop_also = c("race"),      # columns to drop (perhaps 
                                                # other sensitive variable)
                    sample = 0.04,              # sample size from dataset
                    train_size = 0.6,           # size of train set
                    test_size = 0.4,            # size of test set
                    validation_size = 0,        # size of validation set
                    seed = 7                    # seed for reproduction.
)

head(data$train_x,2)
head(data$train_y,2)
head(data$sensitive_train,2)
head(data$test_x,2)
head(data$test_y,2)
head(data$sensitive_test,2)
head(data$data_scaled_test,2)
head(data$data_test,2)
head(data$protected_test,2)

```

We've decided to show the most important objects created by preprocess above.

Our next step is setting a computational device `dev`, which might be GPU with 
cuda or cpu if we don't have cuda installed. Even more importantly we create a 
dataset_loader object which stores data as tensor for our neural network in 4
objects. First two of them are torch `datasets` for storing all the tensors 
and the other two are torch `dataloaders` which store tensors in batches
described by `batch_size`.

```{r}
dev <- if (torch::cuda_is_available()) torch_device("cuda:0") else "cpu"

dsl <- dataset_loader(train_x = data$train_x,
                      train_y = data$train_y,
                      test_x = data$test_x,
                      test_y = data$test_y,
                      batch_size = 5,
                      dev = dev
)
print(dsl$train_dl$.iter()$.next())
```

In the end of preprocessing you can see how the single batch of train data
loader looks like.

## Model creation and pretrain

Finally we are ready to create and pretrain both adversarial and classifier
models. `FairPAN` provides multiple options in this case, because one can not 
only create and pretrain both models with our interface, but also provide their
own neural network models (`clf_model` and `adv_model`). The classifier model 
can be also pretrained, but then `clf_optimizer` from that training must be 
provided and `trained` changed to TRUE.

In our first example we will focus on pretraining both models with creation by
the `pretrain` function. 
To do that you don't have to provide first four
variables, because they are set like that on default. One has to provide `data`
in next four variables and then, the other two describe inner `dataset_loader` 
for adversarial network. Next block describes the structure of our models. Neural
architecture is provided as a simple vector where c(32,16,8) describes a network
with 3 layers with 32, 16 and 8 neurons. This layers are connected with 
`nn_linear` and during forward pass, inner layers have `nnf_relu` with 
`nnf_softmax` in the end. User can also define dimension for softmax, however we
advice not to do so. In the end one can also provide learning rates and numbers
of epochs for both models.
In the end we have to provide a dataset_loader with device and choose whether
we want to monitor more metrics and print them out.

```{r}

models <- pretrain(clf_model = NULL,                       # classifier model
                   adv_model = NULL,                       # adversarial model
                   clf_optimizer = NULL,                   # classifiers optimizer
                   trained = FALSE,                        # indicates whether provided classifier is trained
                   
                   train_x = data$train_x,                 # train predictors
                   train_y = data$train_y,                 # train target
                   sensitive_train = data$sensitive_train, # train sensitives
                   sensitive_test = data$sensitive_test,   # test sensitives
                   
                   batch_size = 5,                         # inner dataset_loader batch size
                   partition = 0.6,                        # partition for inner adversaries dataset_loader preparation
                   
                   neurons_clf = c(32, 32, 32),            # classifiers neural architecture
                   neurons_adv = c(32, 32, 32),            # adversaries neural architecture
                   dimension_clf = 2,                      # dimension for classifier (always set 2)
                   dimension_adv = 1,                      # dimension for adversarial (always set 1)
                   learning_rate_clf = 0.001,              # learning rate of classifier
                   learning_rate_adv = 0.001,              # learning rate of adversarial
                   n_ep_preclf = 5,                        # number of epochs for classifier pretrain
                   n_ep_preadv = 10,                       # number of epochs for adversarial pretrain
                   
                   dsl = dsl,                              # dataset_loader
                   dev = dev,                              # computational device
                   verbose = TRUE,                         # if TRUE prints metrics
                   monitor = TRUE                          # if TRUE aquires more data ( also to print)
)

```
```{r}

clf <- create_model(train_x = data$train_x,                # train predictors
                    train_y = data$train_y,                # train target 
                    neurons = c(32,32,32),                 # models neural architecture
                    dimensions = 2                         # dimension for model (always set 2 for classifier 1 for adversary)
)

opt <- pretrain_net(n_epochs = 5,                          # number of epochs for model pretrain
                    model = clf,                           # neural network model
                    dsl = dsl,                             # dataset_loader
                    model_type = 1,                        # model type (1 means precalssifer)
                    learning_rate = 0.001,                 # learning rate of classifier
                    sensitive_test = data$sensitive_test,  # test sensitives
                    dev = dev,                             # computational device
                    verbose = TRUE,                        # if TRUE prints metrics
                    monitor = TRUE                         # if TRUE aquires more data ( also to print)
)

print(opt$optimizer)

clf_optimizer <- opt$optimizer
    
models <- pretrain(clf_model = clf,                        # classifier model
                   adv_model = NULL,                       # adversarial model
                   clf_optimizer = clf_optimizer,          # classifiers optimizer
                   trained = TRUE,                         # indicates whether provided classifier is trained
                   train_x = data$train_x,                 # train predictors
                   train_y = data$train_y,                 # train target
                   sensitive_train = data$sensitive_train, # train sensitives
                   sensitive_test = data$sensitive_test,   # test sensitives
                   batch_size = 5,                         # inner dataset_loader batch size
                   partition = 0.6,                        # partition for inner adversaries dataset_loader preparation
                   neurons_clf = c(32, 32, 32),            # classifiers neural architecture
                   neurons_adv = c(32, 32, 32),            # adversaries neural architecture
                   dimension_clf = 2,                      # dimension for classifier (always set 2)
                   dimension_adv = 1,                      # dimension for adversarial (always set 1)
                   learning_rate_clf = 0.001,              # learning rate of classifier
                   learning_rate_adv = 0.001,              # learning rate of adversarial
                   n_ep_preclf = 5,                        # number of epochs for classifier pretrain
                   n_ep_preadv = 10,                       # number of epochs for adversarial pretrain
                   dsl = dsl,                              # dataset_loader
                   dev = dev,                              # computational device
                   verbose = TRUE,                         # if TRUE prints metrics
                   monitor = TRUE                          # if TRUE aquires more data ( also to print)
)
```

```{r}
exp_clf <- explain_PAN(target = data$test_y,                     # test target
                       model = models$clf_model,                 # classifier model
                       model_name = "Classifier",                # classifiers name
                       data_test = data$data_test,               # original data for test
                       data_scaled_test = data$data_scaled_test, # scaled numeric data for test
                       batch_size = 5,                           # batch_size used in dataset_loader
                       dev = dev,                                # computational device
                       verbose = TRUE                            #if TRUE prints monitor info
)
```

## Fairtrain

```{r}
monitor <- fair_train( n_ep_pan = 30,                           # number of epochs for pan training
                       dsl = dsl,                               # dataset_loader
                       clf_model = models$clf_model,            # classifier model
                       adv_model = models$adv_model,            # adv model
                       clf_optimizer = models$clf_optimizer,    # classifiers optimizer
                       adv_optimizer = models$adv_optimizer,    # adversaries optimizer
                       dev = dev,                               # computational device
                       sensitive_train = data$sensitive_train,  # train sensitives
                       sensitive_test = data$sensitive_test,    # test sensitives
                       batch_size = 5,                          # inner dataset_loader batch size
                       learning_rate_adv = 0.001,               # learning rate of adversarial
                       learning_rate_clf = 0.001,               # learning rate of classifier
                       lambda = 130,                            # train controlling parameter (the bigger the better STPR results)
                       verbose = TRUE,                          # if TRUE prints metrics
                       monitor = TRUE                           # if TRUE aquires more data ( also to print)
)
monitor
```

## Metrics and Visualizations
```{r}
plot_monitor(STP = monitor$STP, 
             adversary_acc = monitor$adversary_acc, 
             adversary_losses = monitor$adversary_losses, 
             classifier_acc = monitor$classifier_acc)
```


```{r}
exp_PAN <- explain_PAN(target = data$test_y,                     # test target
                       model = models$clf_model,                 # classifier model
                       model_name = "PAN",                       # classifiers name
                       data_test = data$data_test,               # original data for test
                       data_scaled_test = data$data_scaled_test, # scaled numeric data for test
                       batch_size = 5,                           # batch_size used in dataset_loader
                       dev = dev,                                # computational device
                       verbose = TRUE                            #if TRUE prints monitor info
)
DALEX::model_performance(exp_PAN)
DALEX::model_performance(exp_clf)
```

```{r}
fobject <- fairmodels::fairness_check(exp_PAN,exp_clf,
                            protected = data$protected_test,
                            privileged = "Male", 
                            verbose = TRUE)
plot(fobject)
```
